\toolnameend: A Tool for Visualizing Patterns of Spreadsheet Function Combinations

### 2016-08-09 doc/paper.tex
Sentence level correction. Fixes #156.

	From these stories, we can see the dire importance of understanding how people
	actually use spreadsheets. Fortunately, spreadsheet researchers have supported
	this by amassing spreadsheet collections from diverse sources, industrial to
	academic. Collections like EUSES~\cite{fisher2005euses},
	Fuse~\cite{barik2015fuse}, and the Enron corpus~\cite{hermans2015enron} have
	already fueled productive research across the field, such as in
	Hermans'~\cite{hermans2012detecting} and Jansen's~\cite{jansen2015code} search
	for code smells in spreadsheet environments or Aivaloglou and colleague's work
	on making a grammar to capture every possible spreadsheet program in
	Excel~\cite{aivaloglou2015grammar}. Additionally, some research has focused on
	comparing these collections, as Jansen did in 2015, to discover variation in
	spreadsheet techniques over time and location~\cite{jansen2015enron}.

@@

	\section{Related Work} \label{related-work} Visualizations excel for their
	ability to communicate information quickly and
	efficiently~\cite{baeza1999modern}; the visualization of spreadsheets,
	therefore, has offered much room for research. Several previous tools work
	within a spreadsheet, seeking to clarify the connections between the visible
	values in cells, the formulae that produce them, and the dependencies between
	cells that inform them. Igarashi and colleague's fluid
	visualizations~\cite{igarashi1998fluid}, for example, do this by imposing
	lines, color, and animation over the spreadsheet to visually explain these
	tangled connections. Likewise, Clermont explored ways of grouping cells by
	color and border that were related through similar functions, neighbors, or
	references~\cite{clermont2003scalable}, and Hipfl extended this work by
	incorporating the layouts and labels of spreadsheets~\cite{hipfl2008using}.
	Other tools focus on creating visualizations external to the sheets. Hermans
	and colleagues address a spreadsheet programmer's information needs by making
	dataflow diagrams from individual sheets through the tools
	GyroSAT~\cite{hermans2011supporting} and Breviz~\cite{hermans2011breviz}.
	\toolname is among the latter class, creating standalone visualizations about
	spreadsheets instead of drawing connections inside the spreadsheet itself. In
	contrast to the other visualization work, our work focuses on representing functions and how
	they combine into complex formula, rather than exploring the data structures of any
	particular spreadsheet.  \par

@@

	Implicit in this study is an exploration of the Excel API and how people use
	it. Previous research, such as the first study on the Enron
	spreadsheets~\cite{hermans2015enron} or in its comparison with the EUSUS
	corpus~\cite{jansen2015enron}, quantifies how ofter users employ certain
	functions but not how they combine them. Researchers have applied such
	questions to other contexts as well, such as Murphy and colleague's study into
	how Java developers use Eclipse and what commands they use most
	often~\cite{murphy2006java}. Others focus on the documentation of the API
	itself: Robillard and DeLine, for example conducted a series of surveys and
	interviews to diagnose common problems with learning APIs, and they found that
	code examples, one of the focuses of this tool, is one of the most important
	aids to have in learning a new system~\cite{robillard2011field}. Though this
	paper applies the tool to the Enron corpus specifically, the tool itself is not
	limited to this; it can be readily applied to any collection of Excel
	spreadsheets a user gives it.

@@

		\item [!3] \textbf{Do not attempt to explain formulae.} Though the tool will
		to foster understanding by linking pattern to example, it will not try to be
		an explanatory tool that articulates what a complex formula accomplishes. Rather,
		we will provide the structure and context, and the user	must infer this on their own.

@@

	First, a word on the IF function. In Excel, it serves as one of the logical
	functions: given a condition as the first argument that evaluates to either
	true or false, it takes on the value of either the second or third argument,
	respectively. Or, as the documentation writes out, "IF(Something is True, then
	do something, otherwise do something
	else)"\footnote{https://support.office.com/en-us/article/IF-function-69aed7c9-4e8a-4755-a9bc-aa8bbff73be2?ui=en-US\&rs=en-US\&ad=US}. Furthermore, the third argument is optional; if left out, the function will return a default "FALSE" value if necessary. Examples of this function in practice can be seen in Figure~\ref{fig:ifexample}.

@@

	The visualization so far comprises two types of nodes: the circle
	\includegraphics{glossary-greenonly}, which represents a discrete function in
	the formula and is size according to its frequency relative to the number of
	times to root function appeared; and the numbered squares
	\includegraphics{glossary-blue}, which represent the positions of arguments
	within its parent function. From this, we can confirm that, of the times it was
	observed, IF can have at most three arguments passed into it, which corresponds
	with its specifications in the
	API. \par

@@

	We see that IF contains a number of comparison operators as its first argument,
	such as = and $\le$, and boolean-returning functions, like ISNA and AND, with
	simple equality being the most common and some use as ABS being the least seen
	of everything actually used. From here, we can further explore the common options among these functions.
	Clicking on the ``=" node will yield two arguments and expanding each of them
	will peer into the range of common values of equality comparison, which Figure
	\ref{fig:fullpic} shows. \par

@@

	\section{Case Study} The obvious question to pose to any visualization is this:
	What do the images tell us that the text could not? It is not enough to prove
	that something can be visualized; we must also demonstrate what we can be
	gained from any particular visualization. As such, we raised a few tasks in the
	introduction in which we thought the tool could help: detecting bad smells and
	guiding spreadsheet education. To demonstrate the tool's applicability, we
	sought out places in the tree which best exemplify these tasks.

@@

	However, because \toolname makes no judgments beforehand about which designs to
	emphasize as good or bad, the tool can be used to find strange or suboptimal
	designs that exist outside these readily visible categories. Consider the second
	argument of the ``='' operator, already shown expanded in
	Figure~\ref{fig:fullpic}: among the top five argument types for the position in
	an IF function were boolean values. This corresponds to designs in which
	conditions already in a boolean form were then checked against "=TRUE" or
	"=FALSE", such as in the formula \textit{IF(Formulas!C77=TRUE, "Yes")}, which
	appears in one of the sheets. Though this practice does not break anything in
	itself, it nevertheless represents redundant operations in formula design and
	indicates a possible misunderstanding of how the IF function works over the
	hundreds of examples of this boolean comparison that our tool can return.

@@

	%Lookup example
	For a deeper example, the LOOKUP functions are some of the most popular in
	Excel yet come with a number of important but silent requirements for proper
	use, and, as such, they have warranted some extra attention with regards to
	spreadsheet design. In short, these functions are for "when you need to look in
	a single row or column and find a value from the same position in a second row
	or column"\footnote{https://support.office.com/en-us/article/LOOKUP-function-446d94af-663b-451d-8251-369d5e3864cb}. Examples of their use can be found in Figure \ref{fig:lookupexample}, and we will cover each type
	in more depth to better understand the visualizations for these functions.

@@

	All of these variations may seem daunting to a beginner, but \toolname can
	help. Because of its design, users can distinguish between forms of the same
	function, as seen in Figure~\ref{fig:lookups} where the one can explore both
	the frequency with which the spreadsheet programmers used these different forms
	-- three-argument vector form much more than the array form -- and then
	retrieve examples of every configuration in the dataset. The same goes for the
	V/HLOOKUP functions, as shown partially in Figure~\ref{fig:vhlookups}.

@@

	Furthermore, these examples are good as well for exploring how it might bridge
	issues of bad design with education. In a recent study, Hermans, Aivaloglou,
	and Jansen probed the Enron dataset for uses of VLOOKUP and HLOOKUP
	specifically. The issue at hand was the optional fourth argument, which
	requires that rows are sorted lest it return inaccurate results
	\cite{hermans2015detecting}. For someone seeking to increase awareness of this
	problem, it first allows them to answer the question of significance -- "How
	often do spreadsheet programmers use this function, with these particular
	parameters?" Though \toolname does not discern whether neighboring columns are
	sorted, it can nevertheless guide the educator to several examples in
	production where this problem surfaces. \par

@@

	The participants were most vocal about their findings when they discovered, by
	purpose or accident, instances of heinous design in the dataset. Many times,
	they noted these examples shortly after opening a new tree, suggesting that the
	distinctive visual signatures as discussed in Section \ref{badsmells} was a
	salient force in their exploration. For example, when P3 encountered the
	initial design of the SUM tree, as seen in Figure \ref{fig:sum}, they readily
	diagnosed that either a spreadsheet author was "totally incompetent, or [the
	formula was] just something that's weird." Even though the visualization struggles
	to scale well for this formula, it calls clearly to a problem in the underlying
	spreadsheets -- awkward formulae make awkward trees. Other than these visual cues, value
	errors, prefixed with \#s, also appear in the tree, which P4 said could be
	useful for tracking these problems if marked well.

@@

	\subsection{Use in Education} Users found \toolname's inclusion of examples to
	be especially useful since none of the them considered themselves to be experts
	of Excel, their self-reported proficiencies ranging from "fairly familiar" (P2)
	to "not extremely familiar" (P4). For example, as P1 explored the VLOOKUP
	function, they said they could use the tool to collect more examples than the
	documentation offers even though they knew the documentation more verbosely
	explained its function signature. Furthermore, in the process of exploring
	trees, P1 and P3 also gravitated towards some functions they had not used
	before, such as PV and IPMT, looking for new features in Excel to explore and
	learn. \par

>>>>

	From these stories, we can see the dire importance of understanding how people
	actually use spreadsheets. Fortunately, spreadsheet researchers have supported
	this by amassing spreadsheet collections from diverse sources, industrial to
	academic. Collections like EUSES~\cite{fisher2005euses},
	Fuse~\cite{barik2015fuse}, and the Enron corpus~\cite{hermans2015enron} have
	already fueled productive research across the field, such as in the search for
	code smells in spreadsheet environments~\cite{hermans2012detecting}~\cite{jansen2015code} or the creation of grammars to capture every possible
	spreadsheet program in Excel~\cite{aivaloglou2015grammar}. Additionally, some
	research has focused on comparing these collections, as Jansen did in 2015, to
	discover variation in spreadsheet techniques over time and
	location~\cite{jansen2015enron}.

@@

	\section{Related Work} \label{related-work} Visualizations excel for their
	ability to communicate information quickly and
	efficiently~\cite{baeza1999modern}; the visualization of spreadsheets,
	therefore, has offered much room for research. Several previous tools work
	within a spreadsheet, seeking to clarify the connections between the visible
	values in cells, the formulae that produce them, and the dependencies between
	cells that inform them. Igarashi and colleague's fluid
	visualizations~\cite{igarashi1998fluid}, for example, do this by imposing
	lines, color, and animation over the spreadsheet to visually explain these
	tangled connections. Likewise, Clermont explored ways of grouping cells by
	color and border that were related through similar functions, neighbors, or
	references~\cite{clermont2003scalable}, and Hipfl extended this work by
	incorporating the layouts and labels of spreadsheets~\cite{hipfl2008using}.
	Other tools focus on creating visualizations external to the sheets. Hermans
	and colleagues address a spreadsheet programmer's information needs by making
	dataflow diagrams from individual sheets through the tools
	GyroSAT~\cite{hermans2011supporting} and Breviz~\cite{hermans2011breviz}.
	\toolname is among the latter class, creating standalone visualizations about
	spreadsheets instead of drawing connections inside the spreadsheet itself. In
	contrast to the other visualization work, our work focuses on representing functions and how
	they combine into complex formulae, rather than exploring the data structures of any
	particular spreadsheet.  \par

@@

	Implicit in this study is an exploration of the Excel API and how people use
	it. Previous research, such as the first study on the Enron
	spreadsheets~\cite{hermans2015enron} or in its comparison with the EUSUS
	corpus~\cite{jansen2015enron}, quantifies how often users employ certain
	functions but not how they combine them. Researchers have applied such
	questions to other contexts as well, such as Murphy and colleague's study into
	how Java developers use Eclipse and what commands they use most
	often~\cite{murphy2006java}. Others focus on the documentation of the API
	itself: Robillard and DeLine, for example conducted a series of surveys and
	interviews to diagnose common problems with learning APIs, and they found that
	code examples, one of the focuses of this tool, is one of the most important
	aids to have in learning a new system~\cite{robillard2011field}. Though this
	paper applies the tool to the Enron corpus specifically, the tool itself is not
	limited to this; it can be readily applied to any collection of Excel
	spreadsheets a user gives it.

@@

		\item [!3] \textbf{Do not attempt to explain formulae.} Though the tool is meant
		to foster understanding by linking pattern to example, it will not try to be
		an explanatory tool that articulates what a complex formula accomplishes. Rather,
		we will provide the structure and context, and the user	must infer this on their own.

@@

	First, a word on the IF function. In Excel, it serves as one of the logical
	functions: given a condition as the first argument that evaluates to either
	true or false, it takes on the value of either the second or third argument,
	respectively. Or, as the documentation writes out, ``IF(Something is True, then
	do something, otherwise do something
	else)"\footnote{https://support.office.com/en-us/article/IF-function-69aed7c9-4e8a-4755-a9bc-aa8bbff73be2?ui=en-US\&rs=en-US\&ad=US}. Furthermore, the third argument is optional; if left out, the function will return a default ``FALSE" value if necessary. Examples of this function in practice can be seen in Figure~\ref{fig:ifexample}.

@@

	The visualization so far comprises two types of nodes: the circle
	\includegraphics{glossary-greenonly}, which represents a discrete function in
	the formula and is size according to its frequency relative to the number of
	times the root function appeared; and the numbered squares
	\includegraphics{glossary-blue}, which represent the positions of arguments
	within its parent function. From this, we can confirm that, of the times it was
	observed, IF can have at most three arguments passed into it, which corresponds
	with its specifications in the
	API. \par

@@

	We see that IF contains a number of comparison operators as its first argument,
	such as = and $\le$, and boolean-returning functions, like ISNA and AND, with
	simple equality being the most common and ABS being the least. From here, we
	can further explore the common options among these functions. Clicking on the
	``=" node will yield two arguments and expanding each of them
	will peer into the range of common values of equality comparison, which Figure
	\ref{fig:fullpic} shows. \par

@@

	\section{Case Study} The obvious question to pose to any visualization is this:
	What do the images tell us that the text could not? It is not enough to prove
	that something can be visualized; we must also demonstrate what we can
	gain from any particular visualization. As such, we raised a few tasks in the
	introduction in which we thought the tool could help: detecting bad smells and
	guiding spreadsheet education. To demonstrate the tool's applicability, we
	sought out places in the tree which best exemplify these tasks.

@@

	However, because \toolname makes no judgments beforehand about which designs to
	emphasize as good or bad, the tool can be used to find strange or suboptimal
	designs that exist outside these readily visible categories. Consider the second
	argument of the ``='' operator, already shown expanded in
	Figure~\ref{fig:fullpic}: among the top five argument types for the position in
	an IF function were boolean values. This corresponds to designs in which
	conditions already in a boolean form were then checked against ``=TRUE" or
	``=FALSE", such as in the formula \textit{IF(Formulas!C77=TRUE, ``Yes")}, which
	appears in one of the sheets. Though this practice does not break anything in
	itself, it nevertheless represents redundant operations in formula design and
	indicates a possible misunderstanding of how the IF function works over the
	hundreds of examples of this boolean comparison that our tool can return.

@@

	%Lookup example
	For a deeper example, the LOOKUP functions are some of the most popular in
	Excel yet come with a number of important but silent requirements for proper
	use, and, as such, they have warranted some extra attention with regards to
	spreadsheet design. In short, these functions are for ``when you need to look in
	a single row or column and find a value from the same position in a second row
	or column"\footnote{https://support.office.com/en-us/article/LOOKUP-function-446d94af-663b-451d-8251-369d5e3864cb}. Examples of their use can be found in Figure \ref{fig:lookupexample}, and we will cover each type
	in more depth to better understand the visualizations for these functions.

@@

	All of these variations may seem daunting to a beginner, but \toolname can
	help. Because of its design, users can distinguish between forms of the same
	function, as seen in Figure~\ref{fig:lookups} where one can explore both
	the frequency with which the spreadsheet programmers used these different forms
	-- three-argument vector form much more than the array form -- and then
	retrieve examples of every configuration in the dataset. The same goes for the
	V/HLOOKUP functions, as shown partially in Figure~\ref{fig:vhlookups}.

@@

	Furthermore, these examples are good as well for exploring how it might bridge
	issues of bad design with education. In a recent study, Hermans and colleagues
	probed the Enron dataset for uses of VLOOKUP and HLOOKUP
	specifically~\cite{hermans2015detecting}. The issue at hand was the optional
	fourth argument, which requires that rows are sorted lest it return inaccurate
	results. For someone seeking to increase awareness of this problem, it first
	allows them to answer the question of significance -- ``How often do
	spreadsheet programmers use this function, with these particular
	parameters?" Though \toolname does not discern whether neighboring columns are
	sorted, it can nevertheless guide the educator to several examples in
	production where this problem surfaces. \par

@@

	The participants were most vocal about their findings when they discovered, by
	purpose or accident, instances of heinous design in the dataset. Many times,
	they noted these examples shortly after opening a new tree, suggesting that the
	distinctive visual signatures as discussed in Section \ref{badsmells} was a
	salient force in their exploration. For example, when P3 encountered the
	initial design of the SUM tree, as seen in Figure \ref{fig:sum}, they readily
	diagnosed that either a spreadsheet author was ``totally incompetent, or [the
	formula was] just something that's weird." Even though the visualization struggles
	to scale well for this formula, it calls clearly to a problem in the underlying
	spreadsheets -- awkward formulae make awkward trees. Other than these visual cues, value
	errors, prefixed with \#s, also appear in the tree, which P4 said could be
	useful for tracking these problems if marked well.

@@

	\subsection{Use in Education} Users found \toolname's inclusion of examples to
	be especially useful since none of the them considered themselves to be experts
	of Excel, their self-reported proficiencies ranging from ``fairly familiar" (P2)
	to ``not extremely familiar" (P4). For example, as P1 explored the VLOOKUP
	function, they said they could use the tool to collect more examples than the
	documentation offers even though they knew the documentation more verbosely
	explained its function signature. Furthermore, in the process of exploring
	trees, P1 and P3 also gravitated towards some functions they had not used
	before, such as PV and IPMT, looking for new features in Excel to explore and
	learn. \par
